---
title: "CPI FORECAST PROJECT"
author: "Christian"
date: "`r Sys.Date()`"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
```

```{r}
data <- read.csv("C:\\Users\\ENVY\\Downloads\\data1.csv")
head(data, 2)

```

```{r}
# Inspect the structure and first few rows
str(data)
# Summarize the dataset
summary(data)
# Check for missing values
colSums(is.na(data))
```

```{r}
# Step 2: Data Cleaning
# Convert relevant columns to numeric
numeric_columns <- c("Inflation.rates", "CPI", "Production.Tons", 
                     "Monthly_Output_Gap", "total_rainfall", 
                     "average_rainfall", "total_temp_max", "average_temp_max")
data[numeric_columns] <- lapply(data[numeric_columns], function(x) as.numeric(as.character(x)))

# Replace missing numeric values with the mean for each column
data[numeric_columns] <- lapply(data[numeric_columns], function(col) {
  ifelse(is.na(col), mean(col, na.rm = TRUE), col)
})

# Verify if all missing values have been addressed
colSums(is.na(data))

# Check the cleaned data
summary(data)
data

```

```{r}
# Convert the Date column to a Date object with the correct format
data$Date <- as.Date(data$Date, format = "%m/%d/%Y")

# Verify the conversion
head(data$Date)
library(ggplot2)

# Plot CPI trend over time
ggplot(data, aes(x = Date, y = CPI)) +
  geom_line(color = "blue") +
  labs(title = "CPI Trend Over Time", x = "Date", y = "CPI") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


```

```{r}
# Visualize Inflation Rates
ggplot(data, aes(x = Date, y = Inflation.rates)) +
  geom_line(color = "orange") +
  labs(title = "Inflation Rates Over Time", x = "Date", y = "Inflation Rate") +
  theme_minimal()

```

```{r}
# 1. Correlation Analysis
# Determine how strongly CPI and Inflation Rate are correlated with external factors like Production.Tons, total_rainfall, and average_temp_max.
# Compute correlations for numeric variables
numeric_vars <- c("CPI", "Inflation.rates", "Production.Tons", "total_rainfall", "average_temp_max", "Monthly_Output_Gap")
cor_matrix <- cor(data[numeric_vars], use = "complete.obs")

# Visualize the correlation matrix
library(corrplot)
corrplot(cor_matrix, method = "color", addCoef.col = "black", tl.cex = 0.8)

```

# Key Observations from the Correlation Matrix
Strong Correlation Between CPI and Inflation Rates:

CPI and Inflation Rates have a moderate positive correlation (0.55). This is expected since inflation influences CPI.
Negative Correlation Between CPI and Production:

CPI and Production Tons have a negative correlation (-0.35). This could indicate that higher production lowers the cost of goods, affecting the CPI.
Low Correlation with Rainfall and Temperature:

Variables like total rainfall (0.03) and temperature (0.14) have weak correlations with CPI, indicating limited direct impact.
Output Gap and Inflation:

Monthly Output Gap has a weak negative correlation with Inflation Rates (-0.25), suggesting a potential lagged relationship worth exploring.

```{r}
# 1. Feature Selection for Modeling
# Focus on features with higher correlations to CPI, such as Inflation Rates and Production Tons.
# Include lagged variables (e.g., lagged CPI or Inflation Rates) to capture temporal dependencies.
# 2. Lagged Analysis
# Investigate lagged relationships between CPI and other variables, especially Inflation Rates and Output Gap.
# This helps understand delayed effects and can improve forecasting accuracy.
# Create lagged features for CPI and Inflation Rates
data <- data %>%
  mutate(CPI_lag1 = lag(CPI, 1),
         CPI_lag2 = lag(CPI, 2),
         Inflation_lag1 = lag(Inflation.rates, 1),
         Inflation_lag2 = lag(Inflation.rates, 2))

# Check correlation of lagged variables
cor(data[, c("CPI", "CPI_lag1", "CPI_lag2", "Inflation.rates", "Inflation_lag1", "Inflation_lag2")], use = "complete.obs")


```


```{r}
#1. Feature Selection and Engineering:
#Include lagged variables for CPI and inflation rates (CPI_lag1, CPI_lag2, Inflation_lag1, Inflation_lag2) as predictors.
#Add external variables like Production Tons, Rainfall, and Temperature if they are significant
# Select features for the model
features <- c("CPI_lag1", "CPI_lag2", "Inflation_lag1", "Inflation_lag2", "Production.Tons", "total_rainfall", "average_temp_max")
features


```

